---
title: "Biostat 203B Homework 3"
subtitle: Due Mar 1 @ 11:59PM
author: Jonathan Hori - 305947261
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
library(bigrquery)
library(dbplyr)
library(DBI)
library(lubridate)
library(tidyverse)
```

I found the current versions of bigrquery (v1.4.1) and dbplyr (v2.3.0) don't work well together on my computer ([Issue](https://github.com/r-dbi/bigrquery/issues/509)). If you encounter similar issues, you can install older versions of the these two packages.
```{r}
#| eval: false
# library(devtools)
# install_version("bigrquery", version = "1.4.0")
# install_version("dbplyr", version = "2.1.1")
```

## Q1. Compile the ICU cohort in HW2 from the Google BigQuery database 

Below is an outline of steps.

1. Load the GCP BigQuery service account token. Please place the service account token (available at BruinLearn) at your MIMIC data folder: `~/mimic/biostat-203b-2022winter-3fdc2392ac39.json`.
```{r}
# path to the service account token 
satoken <- "~/mimic/biostat-203b-2023winter-3fdc2392ac39.json"
# BigQuery authentication using service account
bq_auth(
  path = satoken,
  # email = "mimiciv-bigquery@biostat-203b-2023winter.iam.gserviceaccount.com",
  # scopes = c("https://www.googleapis.com/auth/bigquery",
  #   "https://www.googleapis.com/auth/cloud-platform")
)
```

1. Connect to BigQuery database `mimic4_v1_0_203b` in GCP (Google Cloud Platform), using the billing account `biostat-203b-2022winter`.
```{r}
# Connect to the BigQuery database `biostat-203b-2022winter.mimic4_v1_0_203b`
con <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2022winter",
    dataset = "mimic4_v1_0_203b",
    billing = "biostat-203b-2022winter"
)
con
```

2. List all tables in the `mimic4_v1_0_203b` database.
```{r}
dbListTables(con)
```

3. Connect to the `icustays` table.
```{r}
# full ICU stays table
icustays_tble <- tbl(con, "icustays") %>%
  show_query() %>%
  print(width = Inf)
```

We only keep the first ICU stay. Following code is kind of a hack, using the `summarise_all(min)` function. It seems that `slice_min()`, `slice_head()`, `distinct(, .keep_all = TRUE)` don't work with `dbplyr`+`bigrquery` at the moment.

```{r}
icustays_tble <- icustays_tble %>% 
  select(subject_id, intime) %>% 
  group_by(subject_id) %>% 
  summarise_all(min) %>%
  left_join(icustays_tble, by = c("subject_id", "intime")) %>%
  show_query() %>%
  print(width = Inf)
```


4. Connect to the `admissions` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id` and `hadm_id`).
```{r}
admissions_tble <- tbl(con, "admissions") %>% 
  semi_join(.,
            icustays_tble,
            by = c("subject_id", "hadm_id")) %>% 
  print(width = Inf)
```

5. Connect to the `patients` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).
```{r}
patients_tble <- tbl(con, "patients") %>% 
  semi_join(.,
            icustays_tble,
            by = "subject_id") %>% 
  print(width = Inf)
```

6. Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW2. 
```{r}
lab_itemid_to_keep <- c(
  50912,
  50971,
  50983,
  50902,
  50882,
  51221,
  51301,
  50931
)


labevents_tble <- tbl(con, "labevents") %>% 
  select(subject_id, itemid, charttime, valuenum) %>% 
  semi_join(.,
            icustays_tble,
            by = "subject_id") %>% 
  filter(itemid %in% lab_itemid_to_keep) %>% 
  print(width = Inf)


# Let's also read the labevents dict:
labitems_dict_tble <- tbl(con, "d_labitems") %>% 
  select(itemid, label) %>% 
  filter(itemid %in% lab_itemid_to_keep) %>% 
  mutate(label_clean = str_to_lower(
    regexp_replace(label, " ", "_"))
  ) %>% 
  select(-label) %>% 
  print(width = Inf)
```
Only keep the first lab measurements during ICU stay and pivot lab items to become variables/columns.
```{r}
labevents_tble <- labevents_tble %>%
  # Join with icustays and filter for time
  left_join(select(icustays_tble, subject_id, intime, outtime),
            by = "subject_id") %>% 
  filter((charttime >= intime) & (charttime <= outtime)) %>% 
  
  # Select first lab measurements
  group_by(subject_id, itemid) %>% 
  arrange(charttime, .by_group = TRUE) %>% 
  filter(row_number() == 1) %>% 
  ungroup() %>%
  
  # Join with itemid dict to get labels
  left_join(labitems_dict_tble, by = "itemid") %>% 
  
  # # Pivot wider
  select(-itemid, -charttime, -intime, -outtime) %>%
  pivot_wider(names_from = label_clean, values_from = valuenum) %>%
  print(width = Inf)
```


7. Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW2.
```{r}
chart_itemid_to_keep <- c(
  220045,
  220181,
  220179,
  223761,
  220210
)

# Import chart events
chartevents_tble <- tbl(con, "chartevents") %>% 
  select(subject_id, charttime, itemid, valuenum) %>% 
  semi_join(.,
            icustays_tble,
            by = "subject_id") %>% 
  filter(itemid %in% chart_itemid_to_keep) %>% 
  print(width = Inf)

# Import itemid dict for labels
chartevents_dict_tble <- tbl(con, "d_items") %>% 
  select(itemid, label) %>% 
  filter(itemid %in% chart_itemid_to_keep) %>% 
  mutate(label_clean = str_to_lower(
    regexp_replace(label, " ", "_"))
  ) %>%
  select(-label) %>% 
  print(width = Inf)
```
Only keep the first chart events during ICU stay and pivot chart events to become variables/columns.
```{r}
chartevents_tble <- chartevents_tble %>% 
  # Filter for intime and outtime
  left_join(select(icustays_tble, subject_id, intime, outtime),
            by = "subject_id") %>% 
  filter((charttime >= intime) & (charttime <= outtime)) %>% 
  
  # Slice first events
  group_by(subject_id, itemid) %>%
  arrange(charttime, .by_group = TRUE) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  
  # Join with labels
  left_join(chartevents_dict_tble, by = "itemid") %>%
  
  # Pivot wider
  select(-itemid, -charttime, -intime, -outtime) %>%
  pivot_wider(names_from = label_clean, values_from = valuenum) %>%
  print(width = Inf)
  
```

8. Put things together. This step is similar to Q7 of HW2. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) start with the `icustays_tble` for the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) create an indicator for 30-day mortality, (vi) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
```

```{r}
icustays_tble %>% 
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) %>% 
  left_join(patients_tble, by = "subject_id") %>% 
  
  # Filter for adults
  mutate(age = year(admittime) - anchor_year + anchor_age) %>% 
  filter(age >= 18) %>% 

  left_join(labevents_tble, by = "subject_id") %>% 
  left_join(chartevents_tble, by = "subject_id") %>% 
  
  # Add thirty day mortality
  mutate(
    # NAs cause problems with casting to integer. First fill with FALSE.
    thirty_day_mort = as.integer(
      ifelse(is.na(deathtime), FALSE,
             sql("(deathtime <= date_add(admittime, interval 30 day)) 
                  and (deathtime > admittime)"))
    )
  ) %>% 
  print(width = Inf) %>% 
  
  # Save to RDS
  collect() %>%
  saveRDS(file = "./mimiciv_shiny/data/icu_cohort.rds")
  
```


Close database connection and clear workspace.
```{r, eval=F}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.

